Software Architectures to Integrate Workflow Engines in Science Gateways

### 2016-05-02 paper.tex
Updates in Sections 2 to 5

Red color is used to represent the components and interactions that
are specific to workflow execution, i.e. that are not involved in any
other process. Tasks refer to computing tasks created by the workflow
engine and executed on the infrastructure. Data represents any type of
file or database involved in workflow execution and stored on the
infrastructure. It does not cover workflow parameters such as strings,
numbers, etc. Administrator refers to a science gateway user with
extended privileges to install applications.

@@

The interactions involved in the architectures are described below and
labeled consistently with Figure~\ref{fig:architectures}.

@@

  the infrastructure, which may require administrative privileges on
  the infrastructure; (\texttt{a$_2$}) the workflow is configured in
  the science gateway so that it becomes available to users. Note that
  integrating a workflow is not the same process as integrating a
  workflow \emph{engine}.

@@

  of the individual tasks, their statuses and so on.

@@

  task, which submitted a sub-task, is usually available and used for

@@

  well-structured and relatively simple workflow language such as between GWorkflowDL

@@

See Figure~\subref*{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a controller
in a Ruby on Rails application. The workflow engine and the science
gateway usually share a database where application, users and other
resources are stored\closednote{sorina}{VIP and Moteur also share the
  WorkflowsDB}{Tristan}{Yes, you are right. This is an example where
  such abstract architectures may not completely apply in real
  life. See comment about ``blended'' architectures in the
  discussion. In case of VIP, the sharing of the DB is more an
  implementation detail rather than a strong architectural pattern: we
  could avoid sharing the DB if we wanted.}.  In this model, task and
data control are both initiated from the science gateway. Interactions
\texttt{b} and \texttt{c$_2$} are initiated from the workflow engine
while \texttt{c$_1$} comes from other parts of the science gateway,
for instance a data management interface. As in any other model, the
installation of new workflows in the science gateway (\texttt{a$_2$})
and infrastructure (\texttt{a$_1$}) is done by an administrator, for
security reasons. This is the model adopted in the Catania Science
Gateway Framework~\cite{Ardizzone2012} (see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}), in the Distributed
application runtime environment
(DARE~\cite{maddineni2012distributed}), in
Galaxy~\cite{goecks2010galaxy}, and in LONI Pipeline
Environment~\cite{dinov2009efficient}. Note that tightly integrated
architectures may provide advanced workflow edition features which are
not covered by our analysis.
%The Catania Science Gateway Framework would be a good illustration here (workflow engine is a Java portlet), but I can't find a suitable figure about it.
%CIPRES~\cite{miller2010creating} (on which the Neuroscience gateway
%described in~\cite{sivagnanam2015early} is based) % \note{Tristan}{There is no workflow engine in there.}

@@

\begin{figure*}
\centering
\def\svgwidth{1.5\columnwidth}
\input{figures/VIP.pdf_tex}
\caption{Architecture of the Virtual Imaging Platform (service
  invocation).  Step \texttt{3} corresponds to interaction \texttt{d}
  on Figure~\ref{fig:architectures}(II), step \texttt{2} corresponds to
  interaction \texttt{c$_1$}, and step \texttt{4} maps to interactions
  \texttt{b} and \texttt{c$_2$} (in VIP, workflow data transfers are
  performed by the tasks and embedded in their descriptions).}
\label{fig:vip-architecture}
\end{figure*}

@@

See Figure~\subref*{archi:sub-task}. The workflow engine is wrapped in
a particular task that can submit sub-tasks to the science gateway
through interaction \texttt{e}. The workflow engine keeps track of the
dependencies between the sub-tasks, but their execution is delegated
to the science gateway that executes them on the infrastructure
through interaction \texttt{b}. Although the science gateway has no
global vision of the workflow, it can keep track of the sub-tasks
submitted by a given task, for instance to be able to cancel them when
the task is canceled. The science gateway may also implement
mechanisms to facilitate the handling of task dependencies, for
instance basic dependency lists as available in most batch managers
(see for instance attribute \texttt{depend} of option \texttt{-W} in
Torque\footnote{\url{http://docs.adaptivecomputing.com}}).

@@

Task encapsulation is implemented in CBRAIN~\cite{SHER-14} where it is
used to integrate the FSL toolkit~\cite{Jenkinson2012782} and the PSOM
workflow engine~\cite{bellec2012pipeline}.

The CBRAIN-FSL integration allows to leverage FSL pipelines written in
low-level workflow languages (Linux executables and scripts) that
submit tasks uniformly through a specific tool called fsl\_sub. It is
diagrammed in Figure~\ref{fig:cbrain-fsl-architecture}, where the
science gateway is represented by components \texttt{CBRAIN portal}
and \texttt{CBRAIN execution server} and the infrastructure consists
of \texttt{Storage servers} and \texttt{Computing cluster with shared
  file system}.

The CBRAIN-PSOM integration~\cite{GLAT-16} is shown on
Figure~\ref{fig:cbrain-psom-architecture}. The PSOM workflow engine
adopts a pilot-job architecture~\cite{turilli2015comprehensive} where
a master coordinates workflow execution by submitting workers and
establishing direct communication channels with them. Note how this
peculiar execution model is well supported by task encapsulation.

\begin{figure*}
\centering
\subfloat[CBRAIN-FSL integration.]{
\def\svgwidth{\columnwidth}
\input{figures/CBRAIN-FSL.pdf_tex}
\label{fig:cbrain-fsl-architecture}
} \hfill \subfloat[CBRAIN-PSOM integration. Figure
reproduced from~\cite{GLAT-16}]{ \def\svgwidth{\columnwidth}
  \input{figures/CBRAIN-PSOM.pdf_tex}
\label{fig:cbrain-psom-architecture}
}
\caption{CBRAIN architecture for workflow engine integration
  (task encapsulation).  1: User sends data and workflow execution request to
  storage server(s) and CBRAIN portal. 2: CBRAIN portal sends
  execution request to execution server on cluster. 3: Execution
  server transfers data from storage server(s). 4-5: Execution server
  starts workflow engine (FSL tool or PSOM master) via task
  scheduler. 6: Workflow engine submits sub-tasks to execution server
  (FSL tasks or PSOM agents). 7-8: Execution server starts sub-tasks
  through task scheduler. FSL sub-tasks will run locally
  instead of being submitted to CBRAIN through interaction 6. 8':
  (PSOM only) PSOM master and workers execute workflow. 9. Execution
  server transfers results to storage server(s). Interaction
  \texttt{b} in Figure~\ref{fig:architectures}(III) is implemented by steps
  \texttt{4}, \texttt{5}, \texttt{7} and \texttt{8} (regular
  interactions with batch manager). Interaction \texttt{c} is
  implemented by steps \texttt{3} and \texttt{9}. Interaction
  \texttt{e} is implemented by step \texttt{6} (for FSL: through a
  modified version of the fsl\_sub script available in
  \url{https://github.com/aces/cbrain-plugins-neuro}; for PSOM:
  through a specific development in PSOM).}
\label{fig:cbrain-sub-tasking}
\end{figure*}

\subsection{Pool model}
\label{sec:pool}
See Figure~\subref*{archi:agent}. Workflows are submitted by the science
gateway to a pool through interaction \texttt{d}. Agents connect to
the pool asynchronously to retrieve and execute workflows through
interaction \texttt{f}. Agents may be started according to various
policies, for instance to ensure load balancing. Agents may wrap
different types of workflow engines. Workflow engine controls tasks
and data on the infrastructure through interactions \texttt{b} and
\texttt{c$_2$}, science gateway transfers user data through
interaction \texttt{c$_1$}, and administrator installs workflows
through interaction \texttt{a$_1$} and \texttt{a$_2$}.

@@

\begin{figure*}
\centering
\includegraphics[width=1.5\columnwidth]{figures/pool-interactions.pdf}
\caption{Architecture of the SHIWA pool. Circle-terminated arrows
  indicate messages that are broadcast to all pool
  clients. Interaction \texttt{d} of Figure~\ref{fig:architectures}(IV) is
  implemented in 3 different types of calls to the pool: workflow
  submission (\texttt{1} \& \texttt{2}), workflow status retrieval
  (\texttt{0} \& \texttt{3})\closednote{sorina}{I suppose letters come
    here from the initial pool-model figure; it's a bit puzzling when
    mixed with letters from Fig 2}{Tristan}{Fixed.}, and workflow
  retrieval (\texttt{13} \& \texttt{14}). \closednote{Marc-e}{Maybe
    easier to follow if the following would be in fig legend instead
    of here, as for 2 previous figures}{Tristan}{fixed} Interaction
  \texttt{f} is implemented through 2 types of calls: workflow
  instance retrieval (\texttt{4}, \texttt{5} and \texttt{6}), and
  workflow instance update (\texttt{11} and \texttt{12}). Workflow
  instance retrieval is used by the agents to fetch work from the
  pool. Workflow instance update is used by the agents to update
  workflow statuses.  Calls \texttt{0}, \texttt{7}, \texttt{8} and
  \texttt{9} are used by
  workflow engine plugins to declare their supported language and to
  launch engines, and by workflow engines to report their status to
  the agent. These calls are specific to the SHIWA Pool implementation
  of the \texttt{agent} component and therefore have no corresponding
  representation in Figure~\ref{fig:architectures}(IV). Figure reproduced
  from~\cite{ROGE-13}.}
\label{fig:shiwa-pool-architecture}
\end{figure*}

@@

Nested workflows are abstract architectural patterns that can be
instantiated in the various architectures described previously. We
focus on instantiation with the service invocation model
(Figure~\subref*{archi:nested}-Right) as this is the most used
architecture. In such an instantiation, we assume that the parent and
child workflow engines are distinct pieces of software that require
different workflow services invoked by distinct \texttt{d}
interactions. If this is not the case, then workflow services can be
collapsed in a single one with a \texttt{d} interaction with
itself. An example of such interaction is the use of hierarchical
workflows in Pegasus~\cite{Deelman201517}.
Workflow engines communicate with infrastructures using
\texttt{b} and \texttt{c$_2$}. Science gateway transfers user data to
infrastructures using \texttt{c$_1$} interactions.

Nested workflows have long been available in workflow engines, for
instance in the Taverna workbench~\cite{oinn2004taverna}. They are
also used implicitly in several platforms where workflow engines are
wrapped in workflow tasks as any other command-line tool. Nested
workflows were notably used by the SHIWA Science Gateway to implement
so-called Coarse-Grained workflow
interoperability~\cite{terstyanszky2014enabling}, i.e. to integrate
various workflow engines in a consistent
platform. Figure~\ref{fig:shiwa-architecture} shows the architecture
used in the SHIWA Simulation Platform for nested workflow execution
with service invocation.
\begin{figure*}
\centering
\includegraphics[width=1.5\columnwidth]{figures/shiwa-science-gateway.pdf}
\caption{Nested workflow execution through SHIWA Science Gateway. The
  parent workflow engine is \texttt{WS-PGRADE}, invoked as a service
  in the Science Gateway (step \texttt{1}, interaction \texttt{d} on
  Figure~\ref{fig:architectures}(V)-Right. Ten different child engines can be used by nested
  workflows, invoked through the \texttt{Submission service} (step
  \texttt{2}, interaction \texttt{d}). Each of these engines can
  submit tasks and transfer data to a distributed computing
  infrastructure (\texttt{DCI}, step \texttt{3}, interactions
  \texttt{b} and \texttt{c$_2$}). Data interactions (\texttt{c}) and
  application porting ones (\texttt{a}) are not represented. Figure
  reproduced from~\cite{terstyanszky2014enabling} with permission of
  the first author.}
\label{fig:shiwa-architecture}
\end{figure*}

@@

See Figure~\subref*{archi:import}. This is an abstract model
instantiated with the service invocation architecture for
consistency. Workflows are integrated in the science gateway through
format conversion from a native format to the science gateway
format. Workflow import is usually an offline process which is not
involved in workflow execution. Workflow import was implemented in the
SHIWA Simulation Platform through the IWIR language that provided a
common language for portability across grid workflow
systems~\cite{plankensteiner-prodan-etal:2013}.  Other systems rely on
workflow import. Tavaxy~\cite{Abouelhoda2012} allows to import, merge
and run Taverna~\cite{oinn2004taverna} and
Galaxy~\cite{goecks2010galaxy} workflows. When some workflow parts
cannot be imported, workflows are run by Tavaxy as nested workflows
using their native engine. The work in~\cite{delaGarza2016} describes
workflow import from KNIME~\cite{Berthold2008} to WS-PGRADE/gUSE and
from Galaxy~\cite{goecks2010galaxy} to WS-PGRADE/gUSE.
\closednote{Tristan}{Add description of a real system}{Tristan}{Would
  require to reproduce a figure from another paper. The concept seems
  simple enough to not require a figure.}

@@

The architectures described in Section~\ref{sec:architectures} are
evaluated in Table~\ref{table:evaluation}. We use 5 main criteria to
evaluate the architectures: integration challenge, execution complexity,
extensibility, scalability and other specific features. Criteria break
down to specific metrics where \emph{lower value indicates better
  performance}. For each criterion, a global score is computed by
summing up the individual metrics. We ensure that the different
metrics in a criterion measure similar entities so that they can be
summed. The criteria and metrics are explained hereafter.

@@

\caption{Architecture evaluation. Lower values (brighter colors) indicate better performance. On each row, metric values are
  normalized between 0 (best value) and 1 (worst value) to determine the
  color of the corresponding table cell. The normalized metric value $m'$ is
  defined as
  $\frac{m-m_{\mathrm{min}}}{m_{\mathrm{min}}-m_{\mathrm{max}}}$ where
  $m$ is the initial metric value, $m_{\mathrm{min}}$ and $m_{\mathrm{max}}$
  are the minimal and maximal values among all architectures. The RGB hexadecimal color code of the cell
  is \texttt{\#99XX99}, where X=\texttt{round}(F-6m') (\texttt{round} rounds a number to the nearest integer). \closednote{Tristan}{Use 6. Round to the nearest integer.}{Tristan}{Fixed.}}
\label{table:evaluation}
\end{table*}

@@

\emph{Extensibility} measures the difficulty to replace or add
elements in the architecture. It is also a consequence of global
complexity and it is determined as the number of interactions and
components that need to be modified when a new element is
added. Modification of a component is required when its code needs to
be updated or recompiled (science gateway or workflow service), or
when a new piece of software has to be installed (infrastructure
only). Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified.  Extensibility
breaks down in 4 metrics depending on the type of element that has to
be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine type (\texttt{E$_1$}): number of interactions or
  components to modify to integrate a new type of workflow engine in
  the architecture. Workflow engines belong to different types when
  they cannot be invoked using the same interface. Adding a new type
  of workflow engine allows to execute more workflows in the science
  gateway.
\item New engine version (\texttt{E$_2$}): number of interactions or
  components to modify to integrate a new version of a workflow engine
  in the architecture, assuming that another version of the same
  engine type is already available. Different versions of a workflow
  engine share the same interface, i.e. they can be invoked using the
  same software. When this is not the case, the different versions are
  considered as different engine types. Components are not treated
  equally regarding engine version updates. Components whose only
  function is to host the workflow engine, i.e. workflow services and
  agents, are not counted in E$_2$ because updates in such components
  are assumed to be straightforward. On the contrary, modifications of
  components that have other functions than wrapping the engine,
  i.e. science gateway and infrastructure, are counted in E$_2$
  because updates in these components require more effort, e.g. new
  release of the science gateway, gaining administrative privileges on
  the infrastructure, etc.
\item New workflow (\texttt{E$_3$}): adding a new workflow is a very
  common operation that does not require modifying software components
  or interactions. We measure the difficulty to integrate a new
  workflow by counting the number of interactions required
  to integrate a new workflow in the architecture, assuming
  that the engine type and version required to execute this workflow
  are already available.
\item New infrastructure (\texttt{E$_4$}): number of interactions or
  components to modify to integrate a new type of infrastructure in
  the architecture. Adding a new infrastructure allows to provide more
  computing or storage power, to access specific types of resources
  (e.g., GPUs, clouds), or to enforce execution policies (e.g.,
  constrain data to remain in a particular network domain).
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The 4 metrics are summed to obtain a global index that measures the
difficulty to extend the architecture. Note that some extensions may
involve several metrics in practice. For instance, adding a new type
of engine may help integrate new infrastructures when interactions
\texttt{b} and \texttt{c$_2$} are already present for the new engine
and infrastructure.  \closednote{Rafael}{Adding a new engine type may
  also allow run in different infrastructures. Maybe it is worth to
  put a sentence to clarify that the metric account for it
  separately.}{Tristan}{It will not *allow* because interaction c1
  would still need to be developed. But it will definitely facilitate
  it. I added a note on this after the description of the metrics.}

>>>>

Red color is used to represent the components and interactions that
are specific to workflow execution, i.e. that are not involved in any
other process. Tasks refer to computing tasks executed on the
infrastructure. Data represents any type of file or database entry involved
in workflow execution and stored on the infrastructure. It does not
cover workflow parameters such as strings, numbers, etc. Administrator
refers to a science gateway user with extended privileges to install
applications.

@@

The interactions involved in the architectures are described below and
labeled as in Figure~\ref{fig:architectures}.

@@

  the infrastructure, which may require administrative privileges;
  (\texttt{a$_2$}) the workflow is configured in the science gateway
  so that it becomes available to users. Note that integrating a
  workflow is not the same process as integrating a workflow
  \emph{engine}.

@@

  of the individual tasks, their statuses, dependencies and so on.

@@

  of a sub-task is usually available and used for

@@

  well-structured and relatively simple workflow languages such as between GWorkflowDL

@@

See Figure~\subref*{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a controller
in a Ruby on Rails application. The workflow engine and the science
gateway usually share a database where application, users and other
resources are stored\closednote{sorina}{VIP and Moteur also share the
  WorkflowsDB}{Tristan}{Yes, you are right. This is an example where
  such abstract architectures may not completely apply in real
  life. See comment about ``blended'' architectures in the
  discussion. In case of VIP, the sharing of the DB is more an
  implementation detail rather than a strong architectural pattern: we
  could avoid sharing the DB if we wanted.}.  In this model, task and
data control are both initiated from the science gateway. Interactions
\texttt{b} and \texttt{c$_2$} are initiated from the workflow engine
while \texttt{c$_1$} comes from other parts of the science gateway,
for instance a data management interface. As in any other model, the
installation of new workflows in the science gateway (\texttt{a$_2$})
and infrastructure (\texttt{a$_1$}) is done by an administrator, for
security reasons. This is the model adopted in the Catania Science
Gateway Framework~\cite{Ardizzone2012} (see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}), in the Distributed
application runtime environment
(DARE~\cite{maddineni2012distributed}), in
Galaxy~\cite{goecks2010galaxy}, and in the LONI Pipeline
Environment~\cite{dinov2009efficient}. Note that tightly integrated
architectures may provide advanced workflow edition features which are
not covered by our analysis.
%The Catania Science Gateway Framework would be a good illustration here (workflow engine is a Java portlet), but I can't find a suitable figure about it.
%CIPRES~\cite{miller2010creating} (on which the Neuroscience gateway
%described in~\cite{sivagnanam2015early} is based) % \note{Tristan}{There is no workflow engine in there.}

@@

\begin{figure*}
\centering
\def\svgwidth{1.5\columnwidth}
\input{figures/VIP.pdf_tex}
\caption{Architecture of the Virtual Imaging Platform (service
  invocation).  Step \texttt{3} corresponds to interaction \texttt{d}
  in Figure~\ref{fig:architectures}(II), step \texttt{2} corresponds to
  interaction \texttt{c$_1$}, step \texttt{4} maps to interaction
  \texttt{b}, and steps 7 and 9 map to \texttt{c$_2$} (in VIP, workflow data transfers are
  performed by the tasks and embedded in their descriptions).}
\label{fig:vip-architecture}
\end{figure*}

@@

See Figure~\subref*{archi:sub-task}. The workflow engine is wrapped in
a particular task that can submit sub-tasks to the science gateway
through interaction \texttt{e}. The workflow engine keeps track of the
dependencies between the sub-tasks, but their execution is delegated
to the science gateway that executes them on the infrastructure
through interaction \texttt{b}. Although the science gateway has no
global vision of the workflow, it can keep track of the sub-tasks
submitted by a given task, for instance to cancel them when
the task is canceled. The science gateway may also implement
mechanisms to facilitate the handling of task dependencies, for
instance basic dependency lists as available in most batch managers
(see for instance attribute \texttt{depend} of option \texttt{-W} in
Torque\footnote{\url{http://docs.adaptivecomputing.com}}).

@@

Task encapsulation is implemented in CBRAIN~\cite{SHER-14} where it is
used to integrate the FSL toolkit~\cite{Jenkinson2012782} and the PSOM
workflow engine~\cite{bellec2012pipeline}. The CBRAIN-FSL integration
allows to leverage FSL pipelines written in low-level workflow
languages (Linux executables and scripts) that submit tasks uniformly
through a specific tool called fsl\_sub. It is diagrammed in
Figure~\subref*{fig:cbrain-fsl-architecture}, where the science gateway is
represented by components \texttt{CBRAIN portal} and \texttt{CBRAIN
  execution server} and the infrastructure consists of \texttt{Storage
  servers} and \texttt{Computing cluster with shared file system}.

The CBRAIN-PSOM integration~\cite{GLAT-16} is shown in
Figure~\subref*{fig:cbrain-psom-architecture}. The PSOM workflow engine
adopts a pilot-job architecture~\cite{turilli2015comprehensive} where
a master coordinates workflow execution by submitting workers and
establishing direct communication channels with them. Note how this
peculiar execution model is well supported by task encapsulation.

\begin{figure*}
\centering
\subfloat[CBRAIN-FSL integration.]{
\def\svgwidth{\columnwidth}
\input{figures/CBRAIN-FSL.pdf_tex}
\label{fig:cbrain-fsl-architecture}
} \hfill \subfloat[CBRAIN-PSOM integration. Figure
reproduced from~\cite{GLAT-16}].{ \def\svgwidth{\columnwidth}
  \input{figures/CBRAIN-PSOM.pdf_tex}
\label{fig:cbrain-psom-architecture}
}
\caption{CBRAIN architecture for workflow engine integration
  (task encapsulation).  1: User sends data and workflow execution request to
  storage server(s) and CBRAIN portal. 2: CBRAIN portal sends
  execution request to execution server on cluster. 3: Execution
  server transfers data from storage server(s). 4-5: Execution server
  starts workflow engine (FSL tool or PSOM master) via task
  scheduler. 6: Workflow engine submits sub-tasks to execution server
  (FSL tasks or PSOM agents). 7-8: Execution server starts sub-tasks
  through task scheduler. FSL sub-tasks will run locally
  instead of being submitted again to CBRAIN through interaction 6. 8':
  (PSOM only) PSOM master and workers execute workflow. 9. Execution
  server transfers results to storage server(s). Interaction
  \texttt{b} in Figure~\ref{fig:architectures}(III) is implemented by steps
  \texttt{4}, \texttt{5}, \texttt{7} and \texttt{8} (regular
  interactions with batch manager). Interaction \texttt{c} is
  implemented by steps \texttt{3} and \texttt{9}. Interaction
  \texttt{e} is implemented by step \texttt{6} (for FSL: through a
  modified version of the fsl\_sub script available in
  \url{https://github.com/aces/cbrain-plugins-neuro}; for PSOM:
  through a specific development in PSOM).}
\label{fig:cbrain-sub-tasking}
\end{figure*}

\subsection{Pool model}
\label{sec:pool}
See Figure~\subref*{archi:agent}. Workflows are submitted by the science
gateway to a central pool through interaction \texttt{d}. Agents connect to
the pool asynchronously to retrieve and execute workflows through
interaction \texttt{f}. Agents may be started according to various
policies, for instance to ensure load balancing. Agents may wrap
different types of workflow engines. Workflow engine controls tasks
and data on the infrastructure through interactions \texttt{b} and
\texttt{c$_2$}, science gateway transfers user data through
interaction \texttt{c$_1$}, and administrator installs workflows
through interaction \texttt{a$_1$} and \texttt{a$_2$}.

@@

\begin{figure*}
\centering
\includegraphics[width=1.5\columnwidth]{figures/pool-interactions.pdf}
\caption{Architecture of the SHIWA pool. Circle-terminated arrows
  indicate messages that are broadcast to all pool
  clients. Interaction \texttt{d} of Figure~\ref{fig:architectures}(IV) is
  implemented by 3 different calls to the pool: workflow
  submission (\texttt{1} \& \texttt{2}), workflow status retrieval
  (\texttt{0} \& \texttt{3})\closednote{sorina}{I suppose letters come
    here from the initial pool-model figure; it's a bit puzzling when
    mixed with letters from Fig 2}{Tristan}{Fixed.}, and workflow
  retrieval (\texttt{13} \& \texttt{14}). \closednote{Marc-e}{Maybe
    easier to follow if the following would be in fig legend instead
    of here, as for 2 previous figures}{Tristan}{fixed} Interaction
  \texttt{f} is implemented through 2 types of calls: workflow
  instance retrieval (\texttt{4}, \texttt{5} and \texttt{6}), and
  workflow instance update (\texttt{11} and \texttt{12}). Workflow
  instance retrieval is used by the agents to fetch work from the
  pool. Workflow instance update is used by the agents to update
  workflow statuses.  Calls \texttt{0}, \texttt{7}, \texttt{8} and
  \texttt{9} are used by
  workflow engine plugins to declare their supported language and to
  launch engines, and by workflow engines to report their status to
  the agent. These calls are specific to the SHIWA Pool implementation
  of the \texttt{agent} component and therefore have no corresponding
  representation in Figure~\ref{fig:architectures}(IV). Figure reproduced
  from~\cite{ROGE-13}.}
\label{fig:shiwa-pool-architecture}
\end{figure*}

@@

Nested workflows are abstract architectural patterns that can be
instantiated in the various architectures described previously. We
focus on instantiation with the service invocation model
(Figure~\subref*{archi:nested}-Right) as this is the most used
architecture. In the instantiation, we assume that the parent and
child workflow engines are distinct pieces of software that require
different workflow services invoked by distinct \texttt{d}
interactions. If this is not the case then workflow services can be
collapsed in a single one with a \texttt{d} interaction with
itself. An example of such interaction is the use of hierarchical
workflows in Pegasus~\cite{Deelman201517}.
Workflow engines communicate with infrastructures using
\texttt{b} and \texttt{c$_2$}. Science gateway transfers user data to
infrastructures using \texttt{c$_1$} interactions.

Nested workflows have long been available in workflow engines, for
instance in the Taverna workbench~\cite{oinn2004taverna}. They are
also used implicitly in several platforms where workflow engines are
wrapped in workflow tasks as any other command-line tool. Nested
workflows were notably used by the SHIWA Science Gateway to implement
so-called Coarse-Grained workflow
interoperability~\cite{terstyanszky2014enabling}, i.e. to integrate
various workflow engines in a consistent
platform. Figure~\ref{fig:shiwa-architecture} shows the architecture
used in the SHIWA Simulation Platform for nested workflow execution
with service invocation.
\begin{figure*}
\centering
\includegraphics[width=1.5\columnwidth]{figures/shiwa-science-gateway.pdf}
\caption{Nested workflow execution through SHIWA Science Gateway. The
  parent workflow engine is \texttt{WS-PGRADE}, invoked as a service
  in the Science Gateway (step \texttt{1}, interaction \texttt{d} on
  Figure~\ref{fig:architectures}(V)-Right). Ten different child engines can be used by nested
  workflows, invoked through the \texttt{Submission service} (step
  \texttt{2}, interaction \texttt{d}). Each of these engines can
  submit tasks and transfer data to a distributed computing
  infrastructure (\texttt{DCI}, step \texttt{3}, interactions
  \texttt{b} and \texttt{c$_2$}). User data interactions (\texttt{c$_1$}) and
  application porting ones (\texttt{a}) are not represented. Figure
  reproduced from~\cite{terstyanszky2014enabling} with permission of
  the first author.}
\label{fig:shiwa-architecture}
\end{figure*}

@@

See Figure~\subref*{archi:import}. This is an abstract model
instantiated with the service invocation architecture for
consistency. Workflows are integrated in the science gateway through
format conversion from a native format to the science gateway
format. Workflow import is usually an offline process that is not
involved in workflow execution. A few systems have implemented
workflow import. In the SHIWA Simulation Platform, it is implemented
through the IWIR language that provides a common language for
portability across grid workflow
systems~\cite{plankensteiner-prodan-etal:2013}. Tavaxy~\cite{Abouelhoda2012}
allows to import, merge and run Taverna~\cite{oinn2004taverna} and
Galaxy~\cite{goecks2010galaxy} workflows; when some workflow parts
cannot be imported, workflows are run by Tavaxy as nested workflows
using their native engine. The work in~\cite{delaGarza2016} describes
workflow import from KNIME~\cite{Berthold2008} to WS-PGRADE/gUSE and
from Galaxy~\cite{goecks2010galaxy} to WS-PGRADE/gUSE.
\closednote{Tristan}{Add description of a real system}{Tristan}{Would
  require to reproduce a figure from another paper. The concept seems
  simple enough to not require a figure.}

@@

The architectures described in Section~\ref{sec:architectures} are
evaluated in Table~\ref{table:evaluation}. We use 5 main criteria to
evaluate the architectures: integration challenge, robustness,
extensibility, scalability and other specific features. Criteria break
down to specific metrics where \emph{lower value indicates better
  performance}. For each criterion, a global score is computed by
summing up the individual metrics. We ensure that the different
metrics in a criterion measure similar entities so that they can be
summed. The criteria and metrics are explained hereafter.

@@

\caption{Architecture evaluation. Lower values (brighter colors) indicate better performance. Cell color is set as follows: (1) on each row, metric values are
  normalized between 0 (best value) and 1 (worst value):
  $m'=\frac{m-m_{\mathrm{min}}}{m_{\mathrm{max}}-m_{\mathrm{min}}}$ where
  $m$ is the metric value, $m_{\mathrm{min}}$ and $m_{\mathrm{max}}$
  are the minimal and maximal values among all architectures; (2) the RGB hexadecimal color code of the cell
  is \texttt{\#99XX99}, where X=\texttt{round}(F-6m') (\texttt{round} rounds a number to the nearest integer). \closednote{Tristan}{Use 6. Round to the nearest integer.}{Tristan}{Fixed.}}
\label{table:evaluation}
\end{table*}

@@

\emph{Extensibility} measures the difficulty to replace or add
elements in the architecture. It is determined as the number of interactions and
components that need to be modified when a new element is
added. Modification of a component is required when its code needs to
be updated or recompiled (science gateway or workflow service), or
when a new piece of software has to be installed (infrastructure
only). Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified.  Extensibility
breaks down in 4 metrics depending on the type of element that has to
be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine type (\texttt{E$_1$}): number of interactions or
  components to modify to integrate a new type of workflow engine in
  the architecture. Workflow engines belong to different types when
  they cannot be invoked using the same interface. Adding a new type
  of workflow engine allows to execute more workflows in the science
  gateway.
\item New engine version (\texttt{E$_2$}): number of interactions or
  components to modify to integrate a new version of a workflow engine
  in the architecture, assuming that another version of the same
  engine type is already available. Different versions of a workflow
  engine share the same interface, i.e. they can be invoked using the
  same software. When this is not the case, the different versions are
  considered as different engine types. Components are not treated
  equally regarding engine version updates. Components whose only
  function is to host the workflow engine, i.e. workflow services and
  agents, are not counted in E$_2$ because updates in such components
  are assumed to be straightforward. On the contrary, modifications of
  components that have other functions than wrapping the engine,
  i.e. science gateway and infrastructure, are counted in E$_2$
  because updates in these components require more effort, e.g. new
  release of the science gateway, gaining administrative privileges on
  the infrastructure, etc.
\item New workflow (\texttt{E$_3$}): adding a new workflow is a very
  common operation that does not require modifying software components
  or interactions. We measure the difficulty to integrate a new
  workflow by counting the number of interactions required
  to integrate a new workflow in the architecture, assuming
  that the engine type and version required to execute this workflow
  are already available.
\item New infrastructure (\texttt{E$_4$}): number of interactions or
  components to modify to integrate a new type of infrastructure in
  the architecture. Adding a new infrastructure allows to provide more
  computing or storage power, to access specific types of resources
  (e.g., GPUs, clouds), or to enforce execution policies (e.g.,
  constrain data to remain in a particular network domain).
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The 4 metrics are summed to obtain a global index that measures the
difficulty to extend the architecture. Note that some extensions may
involve several metrics in practice. For instance, adding a new type
of engine may help integrate new infrastructures when interactions
\texttt{b} and \texttt{c$_2$} are already present for the new engine
and infrastructure.  \closednote{Rafael}{Adding a new engine type may
  also allow run in different infrastructures. Maybe it is worth to
  put a sentence to clarify that the metric account for it
  separately.}{Tristan}{It will not *allow* because interaction c1
  would still need to be developed. But it will definitely facilitate
  it. I added a note on this after the description of the metrics.}
